---
title: "Cell characterisation - merged"
author: "Phil East"
format: html
editor: visual
date: "14/08/2024"
output:
  html_document:
    df_print: tibble
    toc: true
    toc_depth: 4
    toc_float: true
    code_folding: hide
---

```{r}
#| label: setup
#| include: FALSE

knitr::opts_chunk$set(echo=TRUE,warning=FALSE,message=FALSE)
options(cache = TRUE, cache.lazy = FALSE)

```

```{r}
#| label: r_init
#| results: 'hide'
#| warning: FALSE
#| message: FALSE

library( SingleCellExperiment)
library( Spectre )
library( flowCore )
library( tidyverse )
library( DT )

library( flowViz )
library( flowVS )
library( flowAI )
library( PeacoQC )
library( CATALYST )
library( CytoNorm )
library( uwot )
library( grid )
library( ggplot2 )

transformation_home <- file.path( "/project", "analyses", "transformation" )
flowset_file <- file.path( transformation_home, "fs_sub_flowvs.rds" )

script_home <- file.path( "/project", "analyses", "cell-ident-merged" )
sce_file_name <- "merged_sce.rds"
sce_file_path <- file.path( script_home, "objects", sce_file_name )

sce_samp_file_name <- "merged_sce_samp.rds"
sce_samp_file_path <- file.path( script_home, "objects", sce_samp_file_name )
 
ab_file_name <- "antibody_table.csv"
ab_file_path <- file.path( "/project", "data", ab_file_name )
ab_dat <- read.csv( file = ab_file_path )

lineage_markers <- ab_dat %>%
   filter( Lineage ) %>%
   pull( Marker )
sample_n <- 5000

Downsampling_FlowSet <- function(x, samplesize , replace=TRUE, prob=NULL){
  if(missing(samplesize))
    samplesize <- min(flowCore::fsApply(x,nrow))
  flowCore::fsApply(x, function(ff){
    i <- sample(nrow(ff), size = samplesize, replace=replace, prob)
    ff[i,]
  })
}

sce_from_ff <- function( ff, ab_col_ind, ctrl_col_ind, sample_id ) {
  SingleCellExperiment( assays = list( flowvs = t( ff@exprs[, ab_col_ind ] ) ),
                        colData = data.frame( ff@exprs[, ctrl_col_ind ], 
                                              sample_id = sample_id ),
                        rowData =  data.frame( antibody = colnames( ff )[ ab_col_ind ] ),
                        metadata = list( cluster_codes = 1 ) )
}
```

# Merged analysis

We clustered the cells using linear markers to identify broad cell type groups.

```{r}
#| label: merge_samples

fs <- readRDS( flowset_file )

ab_col_ind <- c( 10:42, 44, 45 )
ctrl_col_ind <- c( 1:9, 43, 46 )
antibodies <- colnames( fs )[ ab_col_ind ]

if( file.exists( sce_file_path ) ) {
  sce_merged <- readRDS( sce_file_path )
} else {
  sces <- map( sampleNames( fs ), function( sample_id ) {
    sce_from_ff( fs[[ sample_id ]], 
                 ab_col_ind, 
                 ctrl_col_ind,
                 sample_id )
  } ) %>%
    setNames( sampleNames( fs ) )
  sce_merged <- do.call( 'cbind', sces )
  saveRDS( sce_merged, sce_file_path )
}
```

```{r}
#| label: clustering

if( is.null( metadata( sce_merged )$SOM_codes ) ) {
  set.seed(1234)
  assays( sce_merged )$exprs <- assays( sce_merged )$flowvs
  sce_merged <- cluster( sce_merged, 
                         features=lineage_markers, 
                         maxK=20, 
                         seed=1234 )
  saveRDS( sce_merged, file = sce_file_path )
}
```

```{r}
#| label: subsample

samples <- unique( colData( sce_merged )$sample_id )

if( file.exists( sce_samp_file_path ) ) {
  sce_merged_samp <- readRDS( sce_samp_file_path )
} else {
  set.seed( 1234 )
  sce_merged_samp <- lapply( samples, function( sid ) {
    sce <- sce_merged[, colData( sce_merged )$sample_id == sid ]
    cells <- ncol( sce )
    samp_n <- ifelse( cells < sample_n, cells, sample_n )
    sce[ ,sample( 1:cells, samp_n ) ]
  } ) %>%
    setNames( samples ) %>%
    do.call( 'cbind', . )
  umap_dr <- umap( t( assays( sce_merged_samp )$flowvs[ lineage_markers, ] ), 
                       n_neighbors = 15, 
                       min_dist = 0.1 )
  reducedDim( sce_merged_samp, "UMAP" ) <- umap_dr
  saveRDS( sce_merged_samp, sce_samp_file_path )
}
umap_plot <- plotDR( sce_merged_samp, "UMAP", color_by = "sample_id" )
print( umap_plot )
```

## Clustering

We clustered the subsampled cells using Louvain and SOM clustering

### Louvains

```{r}
#| label: louvains

if( is.null( colData( sce_merged_samp )$louvian ) ) {
  ## Rphenograph
  library(Rphenograph)
  # Assume 'data' is a matrix with cells as rows and markers as columns.
  k <- 200  # You can adjust k based on your dataset
  result <- Rphenograph( t( assays( sce_merged_samp )$flowvs[ lineage_markers, ] ), k )
  clusters <- as.factor(membership(result[[2]]))
  colData( sce_merged_samp )$louvian <- clusters
  saveRDS( sce_merged_samp, sce_samp_file_path )
}

louv_umap <- plotDR( sce_merged_samp, "UMAP", color_by = "louvian", )
print( louv_umap )
```

### Antibody expression

```{r}
#| label: antibody_exprs

ab_map_plots_file <- file.path( script_home, "objects", "merged_ab_umap_plots.rds" )
if( file.exists( ab_map_plots_file ) ) {
    ab_umap_plots <- readRDS( ab_map_plots_file )
} else {
    ab_umap_plots <- rowData( sce_merged_samp )$antibody %>%
	  map( function( i ) {
   	   plotDR( sce_merged_samp, "UMAP", assay = "flowvs", color_by = i )
    	} ) %>%
    setNames( rowData( sce_merged_samp )$antibody )
    saveRDS( ab_umap_plots, file = ab_map_plots_file )
}
```

::: panel-tabset
```{r}
#| label: render_antibody_exprs
#| results: asis
#| echo: false

iwalk( ab_umap_plots, function( gg, label ) { 
  cat('#### ', label, '\n\n')
  grid::grid.newpage()
  grid::grid.draw( gg )
  cat( " \n\n" )
} )
```
:::

```{r}
#| label: facet_cluster
#| fig-width: 12
#| fig-height: 12

louv_umap <- plotDR( sce_merged_samp, "UMAP", color_by = "louvian", facet_by = "louvian" )
print( louv_umap )
```

### FlowSOM

The CATALYST package provides functions to first cluster flow cytometry data with FlowSOM clustering and subsequently perform an UMAP or tSNE with the metacluster labels. Advantage of FlowSOM clustering is the speed of the algorithm and you don’t need to downsample or split your dataset if the exploratory phase is over.

With maxK you specify the number of clusters. The number of clusters to choose can be difficult. First, you need to ask yourself how many clusters would you
expect in your data.You can also use the UMAP of the earlier steps to guide you in choice for number of clusters. Plotting median expression of markers in that UMAP can help to see the number of populations you would expect. Vary the number of clusters to find what best fits your data and is biological relevant.


```{r}
#| label: cluster_plots
#| message: false

cluster_plot_file <- file.path( script_home, "objects", "merged_cluster_plots.rds" )
if( file.exists( cluster_plot_file ) ) {
   cluster_plots <- readRDS( file = cluster_plot_file )
} else {
   meta_clusters <- names( cluster_codes( sce_merged_samp ) )
   cluster_plots <- meta_clusters %>%
      map( function( x ) {
         plotDR( sce_merged_samp, "UMAP", color_by=x )
      } ) %>%
         setNames( meta_clusters )
   saveRDS( cluster_plots, file = cluster_plot_file )
}
```

::: panel-tabset
```{r}
#| label: render_cluster_plots
#| results: asis
#| echo: false

iwalk( cluster_plots, function( gg, label ) { 
  cat('#### ', label, '\n\n')
  grid::grid.newpage()
  grid::grid.draw( gg )
  cat( " \n\n" )
} )
```
:::

```{r}
#| label: cluster_heatmap
#| fig.width: 10
#| fig.height: 12

plotExprHeatmap(sce_merged_samp, features = "type", by="cluster_id", k="meta20", m = "meta4", scale = "last", q = 0, perc=TRUE, bars = TRUE )
```
