---
title: "Preprocess"
author: "Phil East"
format: html
editor: visual
date: "14/08/2024"
output:
  html_document:
    df_print: tibble
    toc: true
    toc_depth: 4
    toc_float: true
    code_folding: hide
---

```{r}
#| label: setup
#| include: FALSE

knitr::opts_chunk$set(echo=TRUE,warning=FALSE,message=FALSE)
options(cache = TRUE, cache.lazy = FALSE)

```

# Preprocess

We read in the spectral flow data and construct SingleCellExperiment objects for each sample. We then assess the individual antibody florecence distributions to determine the best transformation and normalisation.

## FlowCore

### flowSet

`colnames( fs )` returns names of measured covariates per cell (columns in matrix) such as antibodies and autofluorescence channels.

`phenoData( fs )` returns sample metadata table as an AnnotatedDataFrame.

`pData( fs )` returns sample metadata table as a data.frame.

`varLabels( fs )` returns colnames of sample metadata table.

`sampleNames( fs )` returns rownames or sample names from sample metadata table.

### flowFrame

`keyword( fs[[1]] )` returns all parameters associated with each spectral flow sample.

```{r}
#| label: r_init
#| results: 'hide'
#| warning: FALSE
#| message: FALSE

library( SingleCellExperiment)
library( Spectre )
library( flowCore )
library( tidyverse )
library( DT )
library( grid )
library( patchwork )
library( flowTrans )
library( flowViz )
library( flowVS )

jsc <- 'function(settings, json) { $(this).parents(".datatables").remove(); }'

options(bitmapType='cairo')

flow_data_home <- file.path( "/project/data/240904_preprocessed_240903_unmixed")
exp_design_file <- file.path( flow_data_home, "SCT051_sample_list.csv" )
script_home <- file.path( "/project", "analyses", "preprocess" )
```

```{r}
#| include: FALSE
# Init Step to make sure that the dependencies are loaded
htmltools::tagList(datatable(mtcars))
```

## Data In

We read in the FCS files using `flowCore` and create a `flowset` object containg per-sample `flow` data objects containing all cells.

```{r}
#| label: read_fcs

pdat <- read.csv( file = exp_design_file ) %>%
  mutate( sample_id = mouse )
ab_col_ind <- c( 10:42, 44, 45 )

flowset_file <- file.path( script_home, "spectral-flowset.rds" )
if( file.exists( flowset_file )) {
   fs <- readRDS( flowset_file )
} else {
  # Find all FCS in the directory
  fcs_file_paths <- pdat$file %>%
    file.path( flow_data_home, . ) %>%
    setNames( pdat$sample_id )
  fs <- read.flowSet( fcs_file_paths )
  colnames( fs )[ab_col_ind] <- ( fs[[1]]@parameters$desc )[ab_col_ind]
  colnames( fs ) <- gsub( "[ -]", "_", colnames( fs ))
  sampleNames( fs ) <- pdat$sample_id
  saveRDS( fs, flowset_file )
}
```

## Summary stats

We first looked at the summary stats for each of the anitbodies across all samples.

```{r}
#| label: summary_stats

summary_stats <- fs %>%
  map( summary ) %>%
  map( t ) %>%
  map( as.data.frame ) %>%
  map( round ) %>%
  map( rownames_to_column, var = "antibody" ) %>%
  map( DT::datatable ) %>%
  setNames( sampleNames( fs ) )
```

::: panel-tabset
```{r}
#| label: render_summary_stats
#| results: asis
#| echo: false

datatable(matrix(NA, 2, 2), options = list("initComplete" = JS(jsc)))

iwalk( summary_stats, ~ {
    cat(sprintf("\n### %s\n\n", .y))
    print( htmltools::tagList( .x ) )
    cat( ' \n\n')
} )
```
:::

## PnE values

These values encode whether the fluorescent values for a given antibody have been encoded and should linearised when read in. `0,0` indicates no transformation.

```{r}
#| label: pne

keyword( fs[[1]], paste0( "$P", 1:36, "E")) %>%
  as.data.frame( ) %>%
  t( ) %>%
  as.data.frame() %>%
  rownames_to_column( var = "keyword" ) %>%
  mutate( keyword = sub( "X\\.", "", keyword )) %>%
  DT::datatable()
```

## Background metrics

We plotted the distributions of FSC.A, SSC.A and FJComp.AF.A.

```{r}
#| label: bg_metrics

bg_density_plot_file <- file.path( script_home, "bg_density_plots.rds" )
if( file.exists( bg_density_plot_file ) ) {
  bg_density_plots <- readRDS( bg_density_plot_file )
} else {
  bg_density_plots <- fs_sub %>% 
    map( function( ff ) {
      ff@exprs[, c( 1:9, 43, 46 ) ] %>%
        as.data.frame( ) %>% 
        gather( bg_metric, intensity ) %>%
        ggplot( aes( x = intensity, fill = bg_metric, colour = bg_metric )) +
        geom_density( alpha = 0.2 ) +
        labs( x = "bg intensity") +
        scale_x_log10()
    } )
  saveRDS( bg_density_plots, bg_density_plot_file )
}
```

### Distributions

::: panel-tabset
```{r}
#| label: render_bg_dist_plots
#| fig-width: 12
#| fig-cap: backgrond density plots
#| results: asis
#| echo: false

iwalk( bg_density_plots, function( gg, label ) { 
  cat('#### ', label, '\n\n')
  grid::grid.newpage()
  grid::grid.draw( gg )
  cat( " \n\n" )
})
```
:::

## Raw fluoresence distributions

```{r}
#| label: raw_fluor_dists
#| warning: false

raw_density_plot_file <- file.path( script_home, "raw_fluorescence_density_plots.rds" )
if( file.exists( raw_density_plot_file ) ) {
  raw_density_plots <- readRDS( raw_density_plot_file )
} else {
  raw_density_plots <- fs_sub %>% 
    map( function( ff ) {
      ff@exprs[, ab_col_ind ] %>%
        as.data.frame( ) %>%
        gather( antibody, intensity ) %>%
        left_join( antibody_group, by = "antibody" ) %>%
        group_split( group_6 ) %>%
        map( function( df ) {
          ggplot( df, aes( x = intensity, fill = antibody, colour = antibody )) +
          geom_density( alpha = 0.2 ) +
          labs( x = "fluorescence intensity") +
          xlim( -5000, 10000 )
        } )
    } ) %>%
    setNames( sampleNames( fs_sub ) )
  saveRDS( raw_density_plots, raw_density_plot_file )
}
```

### Distributions

::: panel-tabset

```{r}
#| label: render_raw_fluor_dist_plots
#| fig-width: 12
#| fig-height: 18
#| fig-cap: raw fluorescence density plots
#| results: asis
#| echo: false
#| warning: false

iwalk( raw_density_plots, function( ggl, label ) { 
  gg <- ggl |> wrap_plots( ncol = 1, heights = unit( 5, "cm" ))
  cat('#### ', label, '\n\n')
  grid::grid.newpage()
  grid::grid.draw( gg )
  cat( " \n\n" )
})
```
:::
