---
title: "SCT051_spectral"
author: "[Sophie de Carne](mailto:Sophie.deCarne@crick.ac.uk)"
format:
  html:
    embed-resources: true
    code-fold: true
    mainfont: "Avenir"
    css: styles.css
editor: visual
toc: true
toc-depth: 3
toc-expand: 1
toc-location: left
toc_fold: true
anchor-sections: true
---

# Experiment

**Cells and mice \|** KPARG12D cells in C57BL/6J (BRAC). All males because no females with matching age available at the time of the experiment.

**Treatment \|** Started at d-\_. Mice receive AZD4635 (\_mg/kg) \_days a week and aCD73 (250 ug/mouse) IgG control (mIgG2A), A2ARi (AZD4635), anti-CD73 (Oleclumab 2C5 mIgG2A), A2ARi + anti-CD73 combo.

**Cell transplant \|** d0 (14/02/24). 1.5E05 cells injected in 200 ul.

|                      | Luna pre-injection | Luna post-injection |
|----------------------|--------------------|---------------------|
| injected cells, all  | 1.52E05            | 1.09E05             |
| injected cells, live | 1.41E05            | 9.61E04             |
| viability            | 92%                | 88%                 |
| single-cell fraction | 99%                | 100%                |

**Other info \|** Weight is the median of 3 measurements made at d-2, d1 and d5 (12, 15 and 19/02/24).

# Prepare the environment

## Libraries

```{r message=FALSE warning=FALSE}
# data organisation, carpentry
library(plyr)
library(tidyverse) # Suite of packages for data manipulation and visualization (ggplot2, dplyr, tidyr, readr, purrr, tibble)
library(data.table) # Provides an enhanced version of data frames, which are faster and allow easier manipulation of large datasets
library(future.apply)
library(stringr)
library(Rfast)

# to deal with flow data
library(Spectre) # Analysis toolkit for mass cytometry and other high-dimensional data
Spectre::package.load() # Loads necessary dependencies and data from the Spectre package
library(flowCore) # Provides S4 infrastructure for flow cytometry data, including basic data and metadata handling

# dimension reduction
library(Rphenograph) # Implements the PhenoGraph clustering method which is useful for high-dimensional data
library(Rtsne) # Implements T-distributed Stochastic Neighbor Embedding (t-SNE) for dimensionality reduction
library(phateR) # PHATE method for dimensionality reduction, preserving both local and global structures
library(uwot) # Implements the Uniform Manifold Approximation and Projection (UMAP) for dimensionality reduction
library(cytoMEM) # Specifically designed for cytometry data, provides tools for Minimum Event Median dimensionality reduction
library(flowTrans)

# plotting
library(pals) # Provides color palettes for data visualization, enhancing the display of data structures
library(gridExtra) # Provides functions for arranging multiple grid-based plots, enhancing plot layouts
library(ggcyto) # Tailored for visualizing cytometry data using ggplot2 syntax, integrates with the `flowCore` data structures
```

# Load the data

## Select the FCS files to load

```{r}
# Find all FCS in the directory
fileNames <- list.files(
  path = "01_input/FlowJo_preprocessed_files/", pattern = "*.fcs",
  full.names = TRUE, ignore.case = TRUE
)

# Remove the unstained control from the list
unstained_files <- grepl( "unstained", fileNames, ignore.case = TRUE )
if ( length( unstained_files) > 0 ) {
  fileNames <- fileNames[!unstained_files]
}
fileNames
```

## Inspect one file

```{r warning=FALSE}
ff <- read.FCS( fileNames[1], emptyValue = FALSE )
```

### Look at the Flowframe parameters

Description of the parameters (marker names)

```{r}
as.vector( ff@parameters$desc )
```

### Look at the expression

```{r}
View( ff@exprs )
```

## Tidy the flowframe

### Convert flowframe to a datatable as Spectre() works with datatable

```{r}
exprn <- as.data.table( ff@exprs )
```

### Inspect the column names and change to meaningful

```{r}
colnames( exprn )
```

### Rename the column names with `description` (marker names) found in `parameters` in the flowframe

```{r}
colnames( exprn )[4:39] <- ( ff@parameters$desc )[4:39]
setnames( exprn, 3, "AF" )
colnames( exprn )
```

### Add the filename back to the data and add a sample_id

```{r}
exprn <- exprn %>% 
  mutate(
    file = basename( ff@description$FILENAME ),
    sample_id = as.numeric( str_extract( file, "(?<=Well_)[0-9]+" ) )
  ) %>% 
  arrange(
    sample_id
  ) %>% 
  select( sample_id, file, AF, `L_D`, `FSC-A`, `SSC-A`, everything() ) %>% 
  setnames( c( "sample_id", "file", "AF", "L_D", "FSC-A", "SSC-A", "Pd-l1", "Sirpa", "Cd4", "Cd86", "Cd49b", "Cd107a", "B220", "Cd11c", "mKi67", "Tim3", "Cd62l", "Cd3", "Cd44", "Cd11b", "Cd69", "Cd19", "Cd103", "Pd-1", "Ly6g", "Ly6c", "Cd73", "H2-Kb", "Ctla-4", "A2ar", "Nk1.1", "Cd64", "Mertk", "Cd39", "Cd45", "SigF", "Mhc-II", "Cd71", "Cd8a", "Arg1", "Foxp3" ) )

colnames( exprn )
head( exprn )
```

# Build the flowframe with all FCS files

## Load the FCS files into a Flowframe

load the files and add the file names

```{r warning=FALSE}
# Function to process each file
process_file <- function( x ) {
  ff <- read.FCS( x, emptyValue = FALSE )
  this_data <- as.data.table( ff@exprs )
  setnames( this_data, as.vector( ff@parameters$desc ) )  # Set all names at once
  setnames( this_data, 3, "AF" )  # Rename third column to 'AF'
  setnames( this_data, 4:39, ff@parameters$desc[4:39] )  # Rename other columns
  this_data[, file := basename( ff@description$FILENAME )]  # Add file name
  this_data[, sample_id := as.numeric( str_extract( file, "(?<=Well_)[0-9]+" ) )] # extract sample ids
  setorder( this_data, sample_id ) # rank the samples based on sample id
  setcolorder( this_data, c("sample_id", "file", "AF", "L_D" ), before = "FSC-A" )
  return( this_data )
}

# Apply function to all files in parallel
exprn <- lapply( fileNames, process_file )
exprn <- rbindlist( exprn )  # Combine all data.tables
```

## Replace column names with mouse nomenclature

Print the column names of the large datatable

```{r}
colnames( exprn )
```

Rewrite the names

```{r}
exprn <- exprn %>% 
  setnames( c( "sample_id", "file", "AF", "L_D", "FSC-A", "SSC-A", "Pd-l1", "Sirpa", "Cd4", "Cd86", "Cd49b", "Cd107a", "B220", "Cd11c", "mKi67", "Tim3", "Cd62l", "Cd3", "Cd44", "Cd11b", "Cd69", "Cd19", "Cd103", "Pd-1", "Ly6g", "Ly6c", "Cd73", "H2-Kb", "Ctla-4", "A2ar", "Nk1.1", "Cd64", "Mertk", "Cd39", "Cd45", "SigF", "Mhc-II", "Cd71", "Cd8a", "Arg1", "Foxp3" ) )
```

# Transform the data for dimension reduction

## Arcsinus transformation (only run once!)

```{r}
to.asinh <- colnames( exprn )[7:41]
exprn <- do.asinh( exprn, use.cols = to.asinh, cofactor = 10000 )
```

plot the number of events per sample

```{r}
events_per_sample <- table( exprn$sample_id )
barplot( events_per_sample, col = "#0E4A4A" )
```

## Plot the transformed data

Sample the data to save computing time whilst trials and error

```{r}
set.seed( 1234 )
sample_dt <- exprn %>% 
  sample_n( 75000 )
```

Plot the

```{r}
ggplot( exprn, aes( x = Cd4, y = Nk1.1 ) ) +
  geom_hex( bins = 256 ) +
  facet_wrap( ~sample_id )
```
