---
title: "Cell-type Identification"
author: "Phil East"
format: html
editor: visual
date: "14/08/2024"
output:
  html_document:
    df_print: tibble
    toc: true
    toc_depth: 4
    toc_float: true
    code_folding: hide
---

```{r}
#| label: setup
#| include: FALSE

knitr::opts_chunk$set(echo=TRUE,warning=FALSE,message=FALSE)
options(cache = TRUE, cache.lazy = FALSE)

```

```{r}

#| label: r_init
#| results: 'hide'
#| warning: FALSE
#| message: FALSE

library( SingleCellExperiment)
library( Spectre )
library( flowCore )
library( tidyverse )
library( DT )

library( flowViz )
library( flowVS )
library( flowAI )
library( PeacoQC )
library( CATALYST )
library( CytoNorm )
library( uwot )
library( grid )
library( ggplot2 )

flow_data_home <- file.path( "/project/data/240904_preprocessed_240903_unmixed")
exp_design_file <- file.path( flow_data_home, "SCT051_sample_list.csv" )

preprocess_home <- file.path( "/project", "analyses", "preprocess" )
flowset_file <- file.path( preprocess_home, "fs_sub_flowvs.rds" )

script_home <- file.path( "/project", "analyses", "cell-ident" )

Downsampling_FlowSet <- function(x, samplesize , replace=TRUE, prob=NULL){
  if(missing(samplesize))
    samplesize <- min(flowCore::fsApply(x,nrow))
  flowCore::fsApply(x, function(ff){
    i <- sample(nrow(ff), size = samplesize, replace=replace, prob)
    ff[i,]
  })
}
```

# Cell-type Identification

## UMAP

```{r}
#| label: umap

pdat <- read.csv( file = exp_design_file ) %>%
  #mutate( file = file.path( flow_data_home, file )) %>%
  mutate( sample_id = mouse )
fs <- readRDS( flowset_file )

ab_col_ind <- c( 10:42, 44, 45 )

sces <- map( sampleNames( fs ), function( sample_id ) {
  ff <- fs[[ sample_id ]]
  SingleCellExperiment( assays = list( flowvs = t( ff@exprs[, ab_col_ind ] ) ),
                        colData = data.frame( ff@exprs[, c( 1:9, 43, 46 ) ], 
                                              sample_id = sample_id ),
                        rowData =  data.frame( antibody = colnames( ff )[ ab_col_ind ] ),
                        metadata = list( cluster_codes = 1 ) )
} ) %>%
  setNames( sampleNames( fs ) )

sce_merged <- do.call( 'cbind', sces )
```

### Parameter selection

#### n_neighbors

```{r}
#| label: umap_param_nn

set.seed( 1234 )
n_neighbs <- c( 5, 15, 50 )
sce <- sces[[1]]
umap_test_n_neighb <- map( n_neighbs, function( i ) {
  umap_train <- umap( t( assays( sce )$flowvs ), n_neighbors = i )
  reducedDim( sce, "UMAP") <- umap_train
  sce
}) %>%
  setNames( n_neighbs )

umap_plots <- umap_test_n_neighb %>%
  map( plotDR, "UMAP", assay = "flowvs", color_by = "sample_id" )

umap_plots <- names( umap_plots ) %>%
  map( function( i ) {
    umap_plots[[ i ]] +
      labs( title = paste0( "n_neighbours = ", i ) )
  } ) %>%
  setNames( names( umap_plots ) )

print( umap_plots[[1]] )
print( umap_plots[[2]] )
print( umap_plots[[3]] )

```

#### min_dist

```{r}
#| label: umap_param_md

set.seed( 1234 )
min_dist <- c( 0.01, 0.1, 0.5 )
sce <- sces[[1]]
umap_test_min_dist <- map( min_dist, function( i ) {
  umap_train <- umap( t( assays( sce )$flowvs ), n_neighbors = 15, min_dist = i )
  reducedDim( sce, "UMAP") <- umap_train
  sce
}) %>%
  setNames( min_dist )

umap_plots <- umap_test_min_dist %>%
  map( plotDR, "UMAP", assay = "flowvs", color_by = "sample_id" )

umap_plots <- names( umap_plots ) %>%
  map( function( i ) {
    umap_plots[[ i ]] +
      labs( title = paste0( "min_dist = ", i ) )
  } ) %>%
  setNames( names( umap_plots ) )

print( umap_plots[[1]] )
print( umap_plots[[2]] )
print( umap_plots[[3]] )
```

## UMAP merged

```{r}
#| label: umap_merged

umap_merged_file <- file.path( script_home, "sce_merged.rds")
if( file.exists( umap_merged_file ) ) {
  sce_merged <- readRDS( umap_merged_file )
} else {
  set.seed( 1234 )
  umap_merged <- umap( t( assays( sce_merged )$flowvs ), 
                       n_neighbors = 15, 
                       min_dist = 0.1 )
  reducedDim( sce_merged, "UMAP" ) <- umap_merged
  saveRDS( sce_merged, file = sce_merged )
}
umap_plot <- plotDR( sce_merged, "UMAP", assay = "flowvs", color_by = "sample_id")
print( umap_plot )
```

### Antibody expression

```{r}
#| label: antibody_exprs

ab_map_plots_file <- file.path( script_home, "ab_map_plots.rds" )
if( file.exists( ab_map_plots_file ) ) {
    ab_map_plots <- readRDS( ab_map_plots_file )
} else {
    ab_umap_plots <- rowData( sce_merged )$antibody %>%
	  map( function( i ) {
   	   plotDR( sce_merged, "UMAP", assay = "flowvs", color_by = i )
    	} ) %>%
    setNames( rowData( sce_merged )$antibody )
    saveRDS( ab_umap_plots, file = ab_map_plots_file )
}
```

::: panel-tabset
```{r}
#| label: render_antibody_exprs
#| results: asis
#| echo: false

iwalk( ab_umap_plots, function( gg, label ) { 
  cat('#### ', label, '\n\n')
  grid.newpage()
  grid.draw( gg )
  cat( " \n\n" )
} )
```
:::

## Clustering

The CATALYST package provides functions to first cluster flow cytometry data with FlowSOM clustering and subsequently perform an UMAP or tSNE with the metacluster labels. Advantage of FlowSOM clustering is the speed of the algorithm and you don't need to downsample or split your dataset if the exploratory phase is over.

With maxK you specify the number of clusters. The number of clusters to choose can be difficult. First, you need to ask yourself how many clusters would you expect in your data.You can also use the UMAP of the earlier steps to guide you in choice for number of clusters. Plotting median expression of markers in that UMAP can help to see the number of populations you would expect. Vary the number of clusters to find what best fits your data and is biological relevant.

```{r}
#| label: clustering

set.seed(1234)
assays( sce_merged )$exprs <- assays( sce_merged )$flowvs
sce_merged <- cluster( sce_merged, features=NULL, maxK=20, seed=1234 )
saveRDS( sce_merged, file = sce_merged )
```

```{r}
#| label: cluster_plots
#| message: false

meta_clusters <- names( cluster_codes( sce_merged ) )
cluster_plots <- meta_clusters %>%
  map( function( x ) {
    plotDR( sce_merged, "UMAP", color_by=x )
  } ) %>%
  setNames( meta_clusters )
saveRDS( cluster_plots, file = file.path( script_home, "cluster_plots.rds" ) )
```

::: panel-tabset
```{r}
#| label: render_cluster_plots
#| results: asis
#| echo: false

iwalk( cluster_plots, function( gg, label ) { 
  cat('#### ', label, '\n\n')
  grid.newpage()
  grid.draw( gg )
  cat( " \n\n" )
} )
```
:::

#Heatmap of the median expression per marker per metacluster or sample, more information can be found in https://bioconductor.org/packages/release/bioc/html/CATALYST.html

#plotExprHeatmap(sce, features = "type", by="cluster_id",k="meta12",scale = "last", q = 0, perc=TRUE,bars = FALSE)

#plotExprHeatmap(sce, features = "type", by="sample_id",k="meta12",scale = "last", q = 0, perc=TRUE,bars = FALSE) #this plot can also be used to check for batch effects

#Cell_freq_clusters \<- plotAbundances(sce, k = "meta12", group_by = "sample_id") #print(Cell_freq_clusters) #Cell_freq_clusters_data \<- Cell_freq_clusters\[\["data"\]\] #write.xlsx(x=Cell_freq_clusters_data, file="Cellclusterfrequencies.xlsx")

\`\`\`
