---
title: "Transformation"
author: "Phil East"
format: html
editor: visual
date: "14/08/2024"
output:
  html_document:
    df_print: tibble
    toc: true
    toc_depth: 4
    toc_float: true
    code_folding: hide
---

```{r}
#| label: setup
#| include: FALSE

knitr::opts_chunk$set(echo=TRUE,warning=FALSE,message=FALSE)
options(cache = TRUE, cache.lazy = FALSE)

```

# Transformation

```{r}
#| label: r_init
#| results: 'hide'
#| warning: FALSE
#| message: FALSE

library( SingleCellExperiment)
library( Spectre )
library( flowCore )
library( tidyverse )
library( DT )
library( grid )
library( patchwork )
library( flowTrans )
library( flowViz )
library( flowVS )

jsc <- 'function(settings, json) { $(this).parents(".datatables").remove(); }'

options(bitmapType='cairo')

flow_data_home <- file.path( "/project/data/240904_preprocessed_240903_unmixed")
exp_design_file <- file.path( flow_data_home, "SCT051_sample_list.csv" )
script_home <- file.path( "/project", "analyses", "transformation" )
preprocess_home <- file.path( "/project", "analyses", "preprocess" )
```

```{r}
#| include: FALSE
# Init Step to make sure that the dependencies are loaded
htmltools::tagList(datatable(mtcars))
```

## Data In

We read in the FCS files using `flowCore` and create a `flowset` object containg per-sample `flow` data objects containing all cells.

```{r}
#| label: read_fcs
ab_col_ind <- c( 10:42, 44, 45 )
flowset_file <- file.path( preprocess_home, "spectral-flowset.rds" )
if( file.exists( flowset_file )) {
   fs <- readRDS( flowset_file )
}
```

## Subsample cells

We sub sample the cells to create more manageable set of cells (5000). We are using all cells in this analysis.

```{r}
#| label: subsample_cells

all_cells <- TRUE
sample_size <- 5000

if( all_cells ) {
  fs_sub <- fs
} else {
  set.seed( 1000 )
  fs_sub <- fsApply(fs, function(ff) {
    idx <- sample.int(nrow(ff), min(sample_size, nrow(ff)))
    ff[idx,]
  })
}
```

## Transformation

We used flowViz and flowTrans to apply arcsinh transformation to the antibody intensiy values.

-   flowViz applies a dynamic cofactor to the arcsinh transformation

-   flowTrans applies a fixed value.

flowViz's dynamic approach centres the negative background distribution around zero. flowTrans tends to split the negative distribution across zero, depending on the cofactor used (see flowTrans results at the bottom of this report).

```{r}
#| label: transform_markers

markerstotransform <- colnames( fs_sub[[1]] )[ ab_col_ind ]

```

### flowViz arcsin transformation

```{r}
#| label: flowvs

fs_sub_cofactors_file <- file.path( script_home, "fs_sub_cofactors.csv")
if( file.exists( fs_sub_cofactors_file) ) {
  cofactors_df <- read.csv( fs_sub_cofactors_file )
  cofactors <- cofactors_df %>%
    mutate( cofactors = ifelse( !is.na( revised_cofactors), revised_cofactors, cofactors ) ) %>%
    pull( cofactors )
} else {
  cofactors <- estParamFlowVS( fs_sub, channels = markerstotransform )
  cofactors_df <- data.frame( markerstotransform, cofactors )
  write.csv( cofactors_df, file = fs_sub_cofactors_file )
}

fs_sub_flowvs_file <- file.path( script_home, "fs_sub_flowvs.rds")
if( file.exists( fs_sub_flowvs_file) ) {
  fs_sub_flowvs <- readRDS( fs_sub_flowvs_file )
} else {
  fs_sub_flowvs <- transFlowVS( fs_sub, channels = markerstotransform, cofactors )
  saveRDS( fs_sub_flowvs, fs_sub_flowvs_file)
}
cofactors_df %>%
  DT::datatable()
```

### test plots

We generated transformed intensity profiles for antibodies where flowviz was unable to anutomatically identy a suitable arcsinh cofacort. In these instances we selected the cofactor manually.

```{r}
#| label: manual_cofactor
#| eval: false

ab_int <- 10
cofactors[ ab_int ] <- 1000
fs_sub_flowvs <- transFlowVS( fs_sub[1], channels = markerstotransform, cofactors )
gg <- fs_sub_flowvs[[1]]@exprs[, ab_col_ind[ ab_int ] ] %>%
          as.data.frame( ) %>%
          gather( antibody, intensity ) %>%
          ggplot( aes( x = intensity, fill = antibody, colour = antibody )) +
          geom_density( alpha = 0.2 ) +
          labs( x = "asinh fluorescence intensity")
png( file = file.path( script_home, "trans_test.png" ), width = 900 )
print( gg )
dev.off()
```

#### Distributions

```{r}
#| label: asinh_flowvs_dists

bySample <- TRUE

antibody_group <- data.frame( antibody = colnames( fs_sub )[ ab_col_ind ],
                              group_6 = sequence( c( rep( 6, floor( length( ab_col_ind )/6 ) ), length( ab_col_ind ) %% 6 ) ),
                              group = 1:length( ab_col_ind ) )

asinh_flowvs_density_plot_file <- file.path( script_home, "fluorescence_density_plots_asinh_flowvs.rds" )
if( file.exists( asinh_flowvs_density_plot_file ) ) {
  asinh_flowvs_density_plots <- readRDS( asinh_flowvs_density_plot_file )
} else {
  asinh_flowvs_density_plots <- fs_sub_flowvs %>% 
      map( function( ff ) {
        ff@exprs[, ab_col_ind ] %>%
          as.data.frame( ) %>%
          gather( antibody, intensity ) %>%
          left_join( antibody_group, by = "antibody" ) %>%
          group_split( group ) %>%
          map( function( df ) {
            ggplot( df, aes( x = intensity, fill = antibody, colour = antibody )) +
            geom_density( alpha = 0.2 ) +
            labs( x = "asinh fluorescence intensity")
          } ) %>%
             setNames( colnames( ff )[ ab_col_ind ] )
      } ) %>%
        setNames( sampleNames( fs_sub_flowvs ) )
    saveRDS( asinh_flowvs_density_plots, asinh_flowvs_density_plot_file )
}
```

::: panel-tabset
```{r}
#| label: render_asinh_flowvs_fluor_dist_plots
#| fig-width: 9
#| fig-height: 4
#| fig-cap: fluorescence density plots
#| results: asis
#| echo: false

  iwalk( asinh_flowvs_density_plots, function( ggl, label ) { 
    cat('##### ', label, '\n\n')
    cat(':::{.panel-tabset} \n' )
    iwalk( ggl, ~{
      cat('######', .y, '\n\n')
      grid::grid.newpage()
      grid::grid.draw( .x )
      cat( " \n\n" )
    })
    cat(':::\n\n')
  })
```
:::

## Background metrics

We plotted the distributions of FSC.A, SSC.A and FJComp.AF.A.

```{r}
#| label: bg_metrics

bg_density_plot_file <- file.path( script_home, "bg_density_plots.rds" )
if( file.exists( bg_density_plot_file ) ) {
  bg_density_plots <- readRDS( bg_density_plot_file )
} else {
  bg_density_plots <- fs_sub %>% 
    map( function( ff ) {
      ff@exprs[, c( 1:9, 43, 46 ) ] %>%
        as.data.frame( ) %>% 
        gather( bg_metric, intensity ) %>%
        ggplot( aes( x = intensity, fill = bg_metric, colour = bg_metric )) +
        geom_density( alpha = 0.2 ) +
        labs( x = "bg intensity") +
        scale_x_log10()
    } )
  saveRDS( bg_density_plots, bg_density_plot_file )
}
```

### Distributions

::: panel-tabset
```{r}
#| label: render_bg_dist_plots
#| fig-width: 12
#| fig-cap: backgrond density plots
#| results: asis
#| echo: false

iwalk( bg_density_plots, function( gg, label ) { 
  cat('#### ', label, '\n\n')
  grid::grid.newpage()
  grid::grid.draw( gg )
  cat( " \n\n" )
})
```
:::

## Raw fluoresence distributions

```{r}
#| label: raw_fluor_dists
#| warning: false

raw_density_plot_file <- file.path( script_home, "raw_fluorescence_density_plots.rds" )
if( file.exists( raw_density_plot_file ) ) {
  raw_density_plots <- readRDS( raw_density_plot_file )
} else {
  raw_density_plots <- fs_sub %>% 
    map( function( ff ) {
      ff@exprs[, ab_col_ind ] %>%
        as.data.frame( ) %>%
        gather( antibody, intensity ) %>%
        left_join( antibody_group, by = "antibody" ) %>%
        group_split( group_6 ) %>%
        map( function( df ) {
          ggplot( df, aes( x = intensity, fill = antibody, colour = antibody )) +
          geom_density( alpha = 0.2 ) +
          labs( x = "fluorescence intensity") +
          xlim( -5000, 10000 )
        } )
    } ) %>%
    setNames( sampleNames( fs_sub ) )
  saveRDS( raw_density_plots, raw_density_plot_file )
}
```

### Distributions

::: panel-tabset
```{r}
#| label: render_raw_fluor_dist_plots
#| fig-width: 12
#| fig-height: 18
#| fig-cap: raw fluorescence density plots
#| results: asis
#| echo: false
#| warning: false

iwalk( raw_density_plots, function( ggl, label ) { 
  gg <- ggl |> wrap_plots( ncol = 1, heights = unit( 5, "cm" ))
  cat('#### ', label, '\n\n')
  grid::grid.newpage()
  grid::grid.draw( gg )
  cat( " \n\n" )
})
```
:::

### flowTrans

flowTrans gives for following warning for all samples.

```{r}
#| label: flowtrans
#| eval: false
  
fs_sub_trans_file <- file.path( script_home, "fs_sub_trans.rds") 
if( file.exists( fs_sub_trans_file) ) {   
  fs_sub_trans <- readRDS( fs_sub_trans_file ) 
} else {   
  fs_sub_trans <- fsApply( fs_sub, function( x ){     
    flowTrans( x, "mclMultivArcSinh", 
               markerstotransform, 
               n2f=FALSE,              
               parameters.only=FALSE )   
  } )   
  saveRDS( fs_sub_trans, fs_sub_trans_file) 
}
```

### Distributions

```{r}
#| label: asinh_fluor_dists  
#| eval: false

antibody_group <- data.frame( antibody = colnames( fs_sub )[ ab_col_ind ],                                 group = sequence( c( rep( 6, floor( length(ab_col_ind )/6 ) ), length( ab_col_ind ) %% 6 ) ) )  

asinh_density_plot_file <- file.path( script_home, "fluorescence_density_plots_asinh.rds" ) 
if( file.exists( asinh_density_plot_file ) ) {   
  asinh_density_plots <- readRDS( asinh_density_plot_file ) 
} else {   
  asinh_density_plots <- fs_sub_trans %>%      
    map( function( ff ) {       
      ff$result@exprs[, ab_col_ind ] %>%         
        as.data.frame( ) %>%         
        gather( antibody, intensity ) %>%         
        left_join( antibody_group, by = "antibody" ) %>%         
        group_split( group ) %>%         
        map( function( df ) {           
          ggplot( df, aes( x = intensity, fill = antibody, colour = antibody )) +   
            geom_density( alpha = 0.2 ) +           
            labs( x = "asinh fluorescence intensity")         
        } )     
    } )   
  saveRDS( asinh_density_plots, asinh_density_plot_file ) 
}
```

::: panel-tabset
```{r}
#| label: render_asinh_fluor_dist_plots 
#| fig-width: 12 
#| fig-height: 18 
#| fig-cap: fluorescence density plots
#| results: asis 
#| echo: false
#| eval: false

iwalk( asinh_density_plots, function( ggl, label ) {    
  gg <- ggl |> wrap_plots( ncol = 1, heights = unit( 5, "cm" ))   
  cat('#### ', label, '\n\n')   
  grid::grid.newpage()   
  grid::grid.draw( gg )   
  cat( " \n\n" ) 
})
```
:::
