---
title: "Cell QC - BRAC8938.5a"
author: "Phil East"
format: html
editor: visual
date: "14/08/2024"
output:
  html_document:
    df_print: tibble
    toc: true
    toc_depth: 4
    toc_float: true
    code_folding: hide
---

```{r}
#| label: setup
#| include: FALSE

knitr::opts_chunk$set(echo=TRUE,warning=FALSE,message=FALSE)
options(cache = TRUE, cache.lazy = FALSE)

```

```{r}
#| label: r_init
#| results: 'hide'
#| warning: FALSE
#| message: FALSE

library( SingleCellExperiment)
library( Spectre )
library( flowCore )
library( tidyverse )
library( DT )

library( flowViz )
library( flowVS )
library( flowAI )
library( PeacoQC )
library( CATALYST )
library( CytoNorm )
library( uwot )
library( grid )
library( ggplot2 )

sample_id <- "BRAC8938.5a"

#preprocess_home <- file.path( "/project", "analyses", "preprocess" )
#flowset_file <- file.path( preprocess_home, "fs_sub_flowvs.rds" )

transformation_home <- file.path( "/project", "analyses", "transformation" )
flowset_file <- file.path( transformation_home, "fs_sub_flowvs.rds" )

script_home <- file.path( "/project", "analyses", "cell-ident" )
sce_file_name <- paste0( sample_id, "_sce.rds" )
sce_file_path <- file.path( script_home, "objects", sce_file_name )
 
ab_file_name <- "antibody_table.csv"
ab_file_path <- file.path( "/project", "data", ab_file_name )
ab_dat <- read.csv( file = ab_file_path )

lineage_markers <- ab_dat %>%
   filter( Lineage ) %>%
   pull( Marker )

Downsampling_FlowSet <- function(x, samplesize , replace=TRUE, prob=NULL){
  if(missing(samplesize))
    samplesize <- min(flowCore::fsApply(x,nrow))
  flowCore::fsApply(x, function(ff){
    i <- sample(nrow(ff), size = samplesize, replace=replace, prob)
    ff[i,]
  })
}
```

# Cell QC - BRAC8938.5a

## UMAP

```{r}
#| label: sce

fs <- readRDS( flowset_file )

ab_col_ind <- c( 10:42, 44, 45 )

ff <- fs[[ sample_id ]]
sce <- SingleCellExperiment( assays = list( flowvs = t( ff@exprs[, ab_col_ind ] ) ),
                        colData = data.frame( ff@exprs[, c( 1:9, 43, 46 ) ], 
                                              sample_id = sample_id ),
                        rowData =  data.frame( antibody = colnames( ff )[ ab_col_ind ] ),
                        metadata = list( cluster_codes = 1 ) )
```

```{r}
#| label: lineage_markers

lineage_markers <- ab_dat %>%
   filter( Lineage ) %>%
   pull( Marker )
```

## UMAP

```{r}
#| label: umap

if( file.exists( sce_file_path ) ) {
  sce <- readRDS( sce_file_path )
} else {
  set.seed( 1234 )
  umap_dr <- umap( t( assays( sce )$flowvs[ lineage_markers, ] ), 
                       n_neighbors = 15, 
                       min_dist = 0.1 )
  reducedDim( sce, "UMAP" ) <- umap_dr
  saveRDS( sce, file = sce_file_path )
}
```

### Antibody expression

```{r}
#| label: antibody_exprs

ab_map_plots_file <- file.path( script_home, "objects", 
                                paste0( sample_id, "_ab_umap_plots.rds" ) )
if( file.exists( ab_map_plots_file ) ) {
    ab_umap_plots <- readRDS( ab_map_plots_file )
} else {
    ab_umap_plots <- rowData( sce )$antibody %>%
	  map( function( i ) {
   	   plotDR( sce, "UMAP", assay = "flowvs", color_by = i )
    	} ) %>%
    setNames( rowData( sce )$antibody )
    ab_umap_plots$LD <- plotDR( sce, "UMAP", assay = "flowvs", color_by = "Zombie_NIR_A" )
    saveRDS( ab_umap_plots, file = ab_map_plots_file )
}
```

::: panel-tabset
```{r}
#| label: render_antibody_exprs
#| results: asis
#| echo: false

iwalk( ab_umap_plots, function( gg, label ) { 
  cat('#### ', label, '\n\n')
  grid::grid.newpage()
  grid::grid.draw( gg )
  cat( " \n\n" )
} )
```
:::

## Clustering

The CATALYST package provides functions to first cluster flow cytometry data with FlowSOM clustering and subsequently perform an UMAP or tSNE with the metacluster labels. Advantage of FlowSOM clustering is the speed of the algorithm and you don't need to downsample or split your dataset if the exploratory phase is over.

With maxK you specify the number of clusters. The number of clusters to choose can be difficult. First, you need to ask yourself how many clusters would you expect in your data.You can also use the UMAP of the earlier steps to guide you in choice for number of clusters. Plotting median expression of markers in that UMAP can help to see the number of populations you would expect. Vary the number of clusters to find what best fits your data and is biological relevant.

```{r}
#| label: clustering

set.seed(1234)
assays( sce )$exprs <- assays( sce )$flowvs
sce <- cluster( sce, features=lineage_markers, maxK=20, seed=1234 )
saveRDS( sce, file = sce_file_path )
```

```{r}
#| label: cluster_plots
#| message: false

cluster_plot_file <- file.path( script_home, "objects",
                                paste0( sample_id, "cluster_plots.rds" ) )
if( file.exists( cluster_plot_file ) ) {
   cluster_plots <- readRDS( file = cluster_plot_file )
} else {
   meta_clusters <- names( cluster_codes( sce ) )
   cluster_plots <- meta_clusters %>%
      map( function( x ) {
         plotDR( sce, "UMAP", color_by=x )
      } ) %>%
         setNames( meta_clusters )
   saveRDS( cluster_plots, file = cluster_plot_file )
}
```

::: panel-tabset
```{r}
#| label: render_cluster_plots
#| results: asis
#| echo: false

iwalk( cluster_plots, function( gg, label ) { 
  cat('#### ', label, '\n\n')
  grid::grid.newpage()
  grid::grid.draw( gg )
  cat( " \n\n" )
} )
```
:::

```{r}
#| label: cluster_heatmap
#| fig.width: 10
#| fig.height: 12

plotExprHeatmap(sce, features = "type", by="cluster_id", k="meta20", m = "meta4", scale = "last", q = 0, perc=TRUE, bars = TRUE )
```

# cell_freq_clusters \<- plotAbundances(sce, k = "meta12", group_by = "sample_id") #print(Cell_freq_clusters) #Cell_freq_clusters_data \<- Cell_freq_clusters\[\["data"\]\] #write.xlsx(x=Cell_freq_clusters_data, file="Cellclusterfrequencies.xlsx")

\`\`\`
